Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-10-02 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Added refs and text to Chs 2 through 5

The heat flux was measured with water cooled, Schmidt-Boelter total heat flux gauges. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3~\% calibration expanded uncertainty for these devices~\cite{Medtherm}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge was typically 8~\%~\cite{Pitts:2006}.  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values are only presented for the experiments which had two heat flux gauges installed as described in the instrumentation section above.

@@

\subsection{Summary}

@@

\begin{table}
\small
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Measurement     &   \multicolumn{3}{|c|}{Fuels} \\ \cline{2-4}
type                    &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Avg. Peak HRR (kW)      &       80 $\pm$ 9 (11~\%)   &	80 $\pm$ 24 (30~\%) 	& 45 $\pm$ 21 (47~\%)   \\
Total HR (MJ)           &   	23.3 $\pm$ 1.6 (7~\%)   &	13.6 $\pm$ 0.6 (4~\%) 	& 4.1 $\pm$ 0.2 (5~\%)   \\
Avg Peak HF (kW/m$^2$)   &   	2.0 $\pm$ 0.5 (25~\%)   &	3.6 $\pm$ 0.9 (25~\%) 	& 2.5 $\pm$ 0.5 (20~\%)   \\
Median FH (mm)           &   	705 $\pm$ 106 (15~\%)   &	695 $\pm$ 95 (14~\%) 	& 470 $\pm$ 160 (34~\%)   \\
Avg Peak Temp. (C)      &   	410 $\pm$ 103 (25~\%)   &	335 $\pm$ 41 (12~\%) 	& 334 $\pm$ 106 (32~\%)   \\
\hline
\end{tabular}
 \caption[Select measurements for the three source fires]{Select measurements for the three source fires. Each measurement is shown with 95~\% confidence limits both in terms of a value and a percentage.}
 \label{tab:Source_fire_summary}
\end{table}

@@

These instrumented non-combustible wall experiments provide a data set for the fires to be conducted that can be compared against ignition data for gypsum wallboard from the literature to determine the potential for ignition of the paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table~19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate sources.

Quintiere and Harkleroad conducted a study with the Lateral Ignition and Flame Spread (LIFT) apparatus, which they developed ~\cite{ASTM_E1321,Quintiere:1985}.
The ignition temperature value calculated from the LIFT data for regular gypsum wallboard, 12.7~mm thick was 565~$^{\circ}$C with a critical heat flux for ignition of 35~kW/m$^2$.

The second source, was a series of cone calorimeter experiments conducted by Dillon~\cite{ASTM_E1354,Dillon:1998}. The ignition temperature calculated from the cone calorimeter experiments for regular gypsum wallboard, 12.5~mm thick was 515~$^{\circ}$C with a minimum heat flux required for ignition of 26~kW/m$^2$.

A summary of the peak average temperatures and heat fluxes at the non-combustible wall is provided in Table~\ref{tab:TW_Summary}.   Based on the temperature criteria for igniting gyspum wallboard, found in the literature, only the gasoline and polyurethane foam burners positioned against the wall provided temperatures in excess of 515~$^{\circ}$C.  Based on heat flux, each of the burners exceeded the 26~kW/m$^2$ ignition criteria, when positioned against the wall and at when the edge of the burner was 0.15~m away from the wall.

@@

Of course the temperatures of the thermocouples in the wall are a result of the energy transferred to the wall and the thermo-physical properties of the wall material. The properties of the calcium silicate board are simlar but not identical to gypsum wallboard.  The heat flux data is independent of the wall material, so it may be a better predictor of the fire damage to gypsum wallboard when exposed to the burners at different distances.    All of the fuels generated heat fluxes in excess of 26~kW/m$^2$.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.  Based on the heat flux data from Figures~\ref{NCTW_HF_Center_Comp} the natural gas burner should burn the paint and paper off of the face of the gypsum wallboard up to a height of between 0.6~m amd 0.8~m above the burner.  Similarly, the gasoline burner should generate a burn area between 0.8~m and 1.0~m high on a wall composed of gypsum wallboard.  Using the same analysis, the polyurethane foam burner would create a shorter burn region approximately 0.4~m to 0.6~m.  The experiments conducted with the instrumented gypsum wall board will provide data for comparsion with data provided in this chapter and provide insight in to where the may gypsum wall board ignite.

@@

@@

A complete set of data temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set},\ref{IWGB_NG_TC_Backside_set},\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of a an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed.  However at the point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provide distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistant flame region appears to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seem to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.  Note that the thermocouple positioned 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system.

@@

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall the temperatures started at ambient and slowly increased  reaching the peak temperatures at 300~seconds when the burner was shut off.  For the case where the burner was in contact with the wall, the two of the thermocouples located in the persistant flame region indicated a decreasing temperature trend after after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gysum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

@@

The heat flux values from the arrays of sensors mounted flush with the surface of the gypsum wallboard are given in Figures~\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data from the array centered on the burner and the data from the array located at the right edge of the burner are similar in response and magnitude when the burner is positioned away from the wall.  Similar to the wall surface temperatures the heat flux increase as the burner is moved closer to the wall.  However once the burner is in contact with the wall the flame is moving side to side.  As a result, the peaks in the heat flux alternate between the center array and the edge array, as the flame moves and changes shape.  This movement causes changes in the peak heat flux of approximately 20~kW/m$^2$.  The averaged heat flux values graphed with respect to the burner position from the wall are shown in Figure~\ref{IWGB_NG_HF_CandEAvg}.

@@

\subsection{Fire Pattern}

@@

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as shown in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be delineated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are on thermal damage, therefore the focus is on the burned to not burned line of demarcation.

\begin{figure}[p]
	\centering
	\includegraphics[width=2.5in]{../Figures/Soot_pattern}
	\includegraphics[width=2.5in]{../Figures/Soot_close}	\\
	\caption[Photographs showing lines of demarcation with gasoline]{Photographs showing lines of demarcation from a fire pattern generated on painted gypsum wallboard by a gasoline fueled burner.  Burned vs unburned area is clearly defined, where as the difference between sooted and thermally discolored areas is not as clear.}
	\label{Soot_pattern}
\end{figure}

A summary of the peak average temperatures and heat fluxes at the wall is provided in Table~\ref{tab:IWGB_Summary}. While the values in this summary table differ from the values in Table~\ref{tab:TW_Summary} a closer look reveals that most of the values are not ``significantly '' different.  In other words the range of expanded uncertainty values overlap.  For example at 0.0~m position with natural gas, the average peak temperature was 487~$^\circ$C~$\pm$73~$^\circ$C (15~\%). So the upper bound would be 560~$^\circ$C, which includes the corresponding averaged peak value from the instrumented gypsum wallboard experiments of 539~$^\circ$C.  While all of the values do not have overlapping uncertainties, the point is that none of the averaged peak values are ``that '' different.

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath.  The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible was experiments.

@@

Figures~\ref{IWGB_Gas_patterns} and~\ref{IWGB_Gas_0m_patterns} show the progression of thermal damage caused by gasoline fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the gypsum wallboard.  When the burner was positioned at 0.15~m away from the wall, a semi-circular narrow burn line formed with the lowest point approximately 0.1~m above the burner and extending up to the 0.4~m above the burner on the left side.  The formation of the burned area was captured on film and is shown in Figure~\ref{IWGBGAS6_seq}.  The burned region on the gypsum wallboard from exposure to the gasoling fire position 0.08~m away began at 0.2~m above the burner and extended up to 0.6~m. The width of the burned region was approximately one half of the burner width at it's maximum.  Figure~\ref{IWGB_Gas_0m_patterns} shows the fire pattern formed by gas burner when it is tight against the gypsum wallboard.  Two views are shown.  The image on the left shows the wall shortly after the fire went out.  The burned area came be discerned, but a close look is required.  In the photograph on the right, the same piece of gypsum wallboard was lightly brushed off, which revealed a clearer view of the area where the paint and paper were burned off.  The peak height of that fire pattern is just short of 0.8~m. The maxim width is approximately the full width of the burner.

@@

Figures~\ref{IWGB_PUF_patterns} and~\ref{IWGB_PUF_0m_patterns} show the progression of thermal damage caused by the polyurethane fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the instrumented gypsum wallboard.  The burner positioned at 0.15~m from the wall did not transfer enough energy to cause any thermal damage, the only damage to the wall is from soot deposition.  When the burner was located 0.08~m away from the wall, the paint/paper finish over the gypsum was discolored and heavily sooted but the was no consumption or penetration of the paper finish layer.  Three replicate experiments were conducted with the polyurethane foam fueled burner tight against the instrumented gypsum wallboard. The resulting patterns are shown in Figure~\ref{IWGB_PUF_0m_patterns}.  The average height of the burned region of the painted paper finish on the wallboard was 0.3~m above the top of the burner. The maximum width of the burned region was approximately equal to the width of the burner.

@@

With this set of experiments, fire patterns as the result of thermal damage have finally generated in a few cases.  To recap our crude predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  From Chapter III, the gasoline had a potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~ and 0.6~m.  With the polyurethane burner three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatibility of the burn patterns has not been assessed yet, but it appears that the crude predictions from Ch III provided a ``ballpark'' estimate of the maximum height of the burned region.

@@

The purpose of this chapter is to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the small size of the samples needed for the tests.  In the previous sections of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those full scale laboratory experiments provide valuable insight, it would be better to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests in an effort to validate or improve the existing data in the literature.

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption calorimetry method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a conical shaped heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/Cone_Cal} \\
	\caption[Sketch of the cone calorimeter]{Sketch of the cone calorimeter}
	\label{Cone_Cal}
\end{figure}

>>>>

The heat flux was measured with water cooled, Schmidt-Boelter total heat flux gauges. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  An initial series of heat release rate experiments included a heat flux gauge located 2~m from the centerline of the burner and positioned 0.5~m above the top of the burner. the data is provided.  Once the frame was needed to support and position the thermocouples and heat flux gauges, the flux gauge at the 2~m local was not longer used due to concerns fo the frame interferring with the measurement. The manufacturer reports a 3~\% calibration expanded uncertainty for these devices~\cite{Medtherm}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge was typically 8~\%~\cite{Pitts:2006}.  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values in Fig.~\ref{Flux} are only presented for the experiments which had two heat flux gauges installed as described in the instrumentation section above.

In the initial heat release rate experiments another heat flux gauge was used to measure the total heat flux at a location 2~m from the vertical centerline of the burners and 0.5~ above the top of the burners.  At this location the average peak total heat flux measured for the natural gas fires were approximately 0.42~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 0.74~kW/m$^2$ and 0.35~kW/m$^2$ respectively.

@@

@@

\begin{table}
\footnotesize
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Measurement     &   \multicolumn{3}{|c|}{Fuels} \\ \cline{2-4}
type                    &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Peak HRR (kW)      &       80 $\pm$ 9 (11~\%)   &	80 $\pm$ 24 (30~\%) 	& 45 $\pm$ 21 (47~\%)   \\
Total HR (MJ)           &   	23.3 $\pm$ 1.6 (7~\%)   &	13.6 $\pm$ 0.6 (4~\%) 	& 4.1 $\pm$ 0.2 (5~\%)   \\
Peak HF @ 0.95~m (kW/m$^2$)   &   	2.0 $\pm$ 0.5 (25~\%)   &	3.6 $\pm$ 0.9 (25~\%) 	& 2.5 $\pm$ 0.5 (20~\%)   \\
Peak HF @ 2.0~m (kW/m$^2$)   &   	0.42 $\pm$ 0.03 (8~\%)   &	0.74 $\pm$ 0.24 (32~\%) 	& 0.35 $\pm$ 0.18 (51~\%)   \\
Median FH (mm)           &   	705 $\pm$ 106 (15~\%)   &	695 $\pm$ 95 (14~\%) 	& 470 $\pm$ 160 (34~\%)   \\
Peak Temp. (C)      &   	410 $\pm$ 103 (25~\%)   &	335 $\pm$ 41 (12~\%) 	& 334 $\pm$ 106 (32~\%)   \\
\hline
\end{tabular}
 \caption[Select average measurements for the three source fires]{Select average measurements for the three source fires. Each measurement is shown with 95~\% confidence limits both in terms of a value and a percentage.}
 \label{tab:Source_fire_summary}
\end{table}

@@

A series of experiments were conducted with the three different burners positioned a several distances from a non-combustible wall.  The main purpose was to collect data under conditions where the wall would provide a geometric and thermal influence without any contribution to the heat release rate.  A summary of the peak average temperatures and heat fluxes at the face of the non-combustible wall is provided in Table~\ref{tab:TW_Summary}.

@@

This set of experimental results provided an opportunity to make a pre-test fire pattern assessment for the gypsum wallboard experiments which followed.  By comparing the measured temperature and heat flux data against ignition data for gypsum wallboard from the literature, a determination was made on the potential for ignition of the painted, paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table~19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate sources.

Quintiere and Harkleroad conducted a study with the Lateral Ignition and Flame Spread (LIFT) apparatus, a test apparatus which they developed ~\cite{ASTM_E1321,Quintiere:1985}.
The ignition temperature value calculated from the LIFT data for regular gypsum wallboard, 12.7~mm thick was 565~$^{\circ}$C with a critical heat flux for ignition of 35~kW/m$^2$.

The second source, was a series of cone calorimeter experiments conducted by Dillon~\cite{ASTM_E1354,Dillon:1998}. The ignition temperature calculated from the cone calorimeter experiments for regular gypsum wallboard, 12.7~mm thick was 515~$^{\circ}$C with a minimum heat flux required for ignition of 26~kW/m$^2$.

Of course the temperatures of the thermocouples in the wall are a result of the energy transferred to the wall and the thermo-physical properties of the wall material. The properties of the calcium silicate board are simlar but not identical to gypsum wallboard.  The heat flux data is independent of the wall material, so it may be a better predictor of the fire damage to gypsum wallboard when exposed to the burners at different distances.    All of the fuels generated heat fluxes in excess of 26~kW/m$^2$.  However only the burner against the wall positions provided a combination of heat flux and temperature that had the potential to ignite the paper on the gypsum wallboard based on the handbook data.  Based on the heat flux data from Figures~\ref{NCTW_HF_Center_Comp} the natural gas burner should burn the paint and paper off of the face of the gypsum wallboard up to a height of between 0.6~m amd 0.8~m above the burner.  Similarly, the gasoline burner should generate a burn area between 0.8~m and 1.0~m high on a wall composed of gypsum wallboard.  Using the same analysis, the polyurethane foam burner would create a shorter burn region approximately 0.4~m to 0.6~m.  The experiments conducted with the instrumented painted gypsum wallboard will provide data for comparsion with data provided in this chapter and provide insight in to where the gypsum wallboard ignited and if the paint and paper contributed to the heat release rate and spread of the fire pattern.

@@

McGraw and Mowrer conducted a study to examine both the flammability and dehydration of unpainted and painted gypsum wallboard~\cite{McGraw:1999}.  Their study exposed gypsum wallboard samples to heat fluxes of 25, 50, and 75~kW/m~$^2$ using a cone calorimeter.  None of the samples ignited at 25~kW/m~$^2$, even with exposure times of 15 minutes. The results showed that two coats of latex paint had little effect on the time to ignition at 50~kW/m~$^2$.  However it did increase the peak heat release rate per unit area from 111~kW/m~$^2$ to 211~kW/m~$^2$.  The total heat released per unit area over the course of the exposure also increased from 1561~kJ/m~$^2$ to 2284~kJ/m~$^2$. After the Type X gypsum wallboard samples were exposed to the heat flux, the samples were cut and the depth of dehydration was assessed by length measurement.  Analytic means were used to determine that two coats of latex paint would not contribute the flame spread on the gypsum wallboard and using the moisture mass loss, a secondary means was used to approximate the depth of calcination.  The limited impact of the paint on the flammability of the gypsum board was important in the development of this current study.

@@

A complete set of data temperature and heat flux data for the natural gas burner experiments conducted at 0.31~m, 0.15~m, and 0.0~m are provided in Figures~\ref{IWGB_NG_TC_Plume_set},\ref{IWGB_NG_TC_Wall_set},\ref{IWGB_NG_TC_Backside_set},\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data for the burner positioned 0.31~m from the wall is the average of three replicate experiments.  Note that the data set for the burner position 0.0~m from the wall only has 180~seconds of fire data.  This was the result of an early shut down of the gas supply.  This decrepancy was not noticed until the data was reviewed.  However at the point there was no opportunity to rerun the experiment.

The temperatures measured along the vertical center axis of the burner provide distinct differences between the flame regions and the buoyant plume region based on the significant differences in temperature see Figure~\ref{IWGB_NG_TC_Plume_set}.  The persistant flame region appears to be recorded at the thermocouple positions 0.2~ through 0.6~m above the burner.  Based on the temperature time histories, the thermocouples positioned 0.8~m through 1.2~m seem to be in the intermittent flame region.  The thermocouples position at or above 1.4~m appear to be in the buoyant plume region.  The fluctuations in the peak temperatures for this ``steady state'' fire are the result of the flame movement away from the thermocouples centered above the burner.

@@

The temperatures measured on the front surface of the painted gypsum wallboard are displayed in Figure~\ref{IWGB_NG_TC_Wall_set}.  When the burner was located 0.31~m and 0.15~m away from the wall the temperatures started at ambient and slowly increased  reaching the peak temperatures at 300~seconds when the burner was shut off.  Note that during the 0.15~m burner position experiment, the thermocouple located 0.2~m above the burner did not increase in temperature due to a connection problem in the data acquisition system. For the case where the burner was in contact with the wall, the two of the thermocouples located in the persistant flame region indicated a decreasing temperature trend after after the paper on the gypsum wallboard had burned away.  The decrease in temperature my have had been the result of material from the burned gysum wallboard shielding otherwise impacting those thermocouples. The thermocouple positioned 0.2~m above the burner was exhibiting the same behavior as in the previous experiment, but the problem was resolved during the experiment.  The restored connection to the data system accounts for the rapid rise in temperature at 60~seconds into the experiment.

@@

The heat flux values from the arrays of sensors mounted flush with the surface of the gypsum wallboard are given in Figures~\ref{IWGB_NG_HF_CenterSet}, and \ref{IWGB_NG_HF_EdgeSet}.  The data from the array centered on the burner and the data from the array located at the right edge of the burner are similar in response and magnitude when the burner is positioned away from the wall.  Similar to the wall surface temperatures the heat flux increase as the burner is moved closer to the wall.  However once the burner is in contact with the wall the flame is moving side to side.  As a result, the peaks in the heat flux alternate between the center array and the edge array, as the flame moves and changes shape.  This movement causes changes in the peak heat flux of approximately 20~kW/m$^2$.

In Figure~\ref{IWGB_NG_HF_CandEAvg}, the averaged peak heat flux values for the natural gas burner experiments are graphed with respect to the distance that the burner was positioned from the gypsum wallboard. The data for both of the vertical heat flux gauge arrays, located on the centerline and the right edge of the burner are provided. In general the heat flux trends from higher in the positions near the wall to lower as the burner was positioned further from the wall.  However once the burner was within 0.15~m of the wall or less, the flame was drawn toward the wall.  As a result some of the peak heat fluxes occurred at heights that were out of sync with the trends at the positions further from the wall.

@@

\subsection{Fire Patterns}

@@

Soot and discoloration alone were not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as shown in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be delineated while the discolored and sooted areas tended to blend together.  Further the focus of the measurements in this study are on thermal damage, therefore the focus is on the ``burned to not burned'' line of demarcation.

\begin{figure}[p]
	\centering
	\includegraphics[width=2.5in]{../Figures/Soot_pattern}
	\includegraphics[width=2.5in]{../Figures/Soot_close}	\\
	\caption[Photographs showing lines of demarcation with gasoline]{Photographs showing lines of demarcation from a fire pattern generated on painted gypsum wallboard by a gasoline fueled burner.  Burned vs unburned area was clearly defined, where as the difference between sooted and thermally discolored areas was not as clear.}
	\label{Soot_pattern}
\end{figure}

Figure~\ref{IWGB_NG_patterns} provides a comparison between the area of discoloration of the paint with the natural gas burner positioned at 0.15~m away from the wall versus the area of consumed and penetrated paint/paper layer which exposed the gypsum underneath.  The peak height of the consumed and penetrated paper, i.e. the burned region, was 0.8~above the burner.  This is consistent with the prediction based on the lower heat flux ignition value found in the literature and the average peak heat flux from the non-combustible wall experiments.

@@

Figures~\ref{IWGB_Gas_patterns} and~\ref{IWGB_Gas_0m_patterns} show the progression of thermal damage caused by gasoline fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the gypsum wallboard.  When the burner was positioned at 0.15~m away from the wall, a semi-circular narrow burn line formed with the lowest point approximately 0.1~m above the burner and extending up to the 0.4~m above the burner on the left side.  The formation of the burned area was captured on film and is shown in Figure~\ref{IWGBGAS6_seq}.  The burned region on the gypsum wallboard from exposure to the gasoline fire position 0.08~m away began at 0.2~m above the burner and extended up to 0.6~m. The width of the burned region was approximately one half of the burner width at it's maximum.  Figure~\ref{IWGB_Gas_0m_patterns} shows the fire pattern formed by gas burner when it was tight against the gypsum wallboard.  Two views are shown.  The image on the left shows the wall shortly after the fire went out.  The burned area can be discerned, but a close look is required.  In the photograph on the right, the same piece of gypsum wallboard was lightly brushed off, which revealed a clearer view of the area where the paint and paper were burned off.  The peak height of that fire pattern is just short of 0.8~m. The maxim width is approximately the full width of the burner.

@@

Figures~\ref{IWGB_PUF_patterns} and~\ref{IWGB_PUF_0m_patterns} show the progression of thermal damage caused by the polyurethane fueled burner at 0.15~m, 0.08~m, and 0.0~m away from the instrumented gypsum wallboard.  The burner positioned at 0.15~m from the wall did not transfer enough energy to cause any thermal damage, the only damage to the wall is from soot deposition.  When the burner was located 0.08~m away from the wall, the paint/paper finish over the gypsum was discolored and heavily sooted but there was no consumption or penetration of the paper finish layer.  Three replicate experiments were conducted with the polyurethane foam fueled burner tight against the instrumented gypsum wallboard. The resulting patterns are shown in Figure~\ref{IWGB_PUF_0m_patterns}.  The average height of the burned region of the painted paper finish on the wallboard was 0.3~m above the top of the burner. The maximum width of the burned region was approximately equal to the width of the burner.  The pattern generated from the polyurethane burner was different in shape and composition when compared to the patterns generated by the natural gas and gasoline burners.  The shape of the polyurethane generated pattern was triangular with the base near the top edge of the burner and the peak toward the vertical centerline of the burner.  The natural gas and gasoline patterns were much taller and typically had a region just above the top edge of the burner that narrowed and then widened as the height of the pattern increased.  The patterns reached a near steady width, 200 to 300~mm above the burner and then narrowed to a point near the top of the fire pattern.  Another difference had to do with handling the patterns post test.  While the the loose soot and burnt paper could be brushed away with a ``light touch'' for the brush, that was not the case for the polyurethane patterns.  The burnt paper was still attached to the gypsum and the soot seemed to adhere to the gypsum as well.  As a result, post test brushing the patterns would not be done conducted for the compartment wall experiments.  Among the concerns was an introduction of other uncertainties with regard to: how hard the brush was applied, what was the bristle stiffness of the brush, and would a new brush be needed each time?

@@

A summary of the peak average temperatures and heat fluxes measured at the wall surface is provided in Table~\ref{tab:IWGB_Summary}. While the values in this summary table differ from the values in Table~\ref{tab:TW_Summary} a closer look reveals that most of the values are not ``significantly '' different.  In other words the range of expanded uncertainty values overlap.  For example at 0.0~m position with natural gas, the average peak temperature was 487~$^\circ$C~$\pm$73~$^\circ$C (15~\%). So the upper bound would be 560~$^\circ$C, which includes the corresponding averaged peak value from the instrumented gypsum wallboard experiments of 539~$^\circ$C.  While a few of the values do not have overlapping uncertainties, the point is that none of the averaged peak values are ``that '' significantly different.

With this set of experiments, fire patterns as the result of thermal damage were finally generated in a few cases.  To recap our predictions from the previous chapter, the natural gas burner had the potential to cause a fire pattern with a height between 0.6~m and 0.8~m.  With the natural gas burner against the wall, a pattern with a maximum burn area of 0.8~m in height was generated.  Also from Chapter 3, according to the heat flux and the ignition data from the literature, the gasoline had a potential to create a fire pattern with a maximum height between 0.8~m and 1.0~m.  In the single instrumented gypsum wallboard experiment with the gasoline burner located against the wall, the maximum height of the burned region was just short of 0.8~m.  The prediction for the potential height of the fire pattern that could be created with the polyurethane burner against the wall was between 0.4~ and 0.6~m.  With the polyurethane burner three replicate experiments were conducted.  The range of the maximum heights of the burned region was 0.2~m to 0.35~m.  The average peak burned region height was 0.3~m. The uncertainty of the repeatibility of the burn patterns has not been assessed yet, but it appears that the predictions from Ch 3 provided reasonable ``ballpark'' estimates of the maximum height of the burned region.

@@

The purpose of this chapter was to develop and compare data from bench scale apparatus that could be used to analyze fire damage to gypsum wallboard.
This type of testing might be practical for a fire investigator to use based on the low cost, relative to a full-scale fire test, and the relatively small size of the samples needed for the tests.  In the previous chapters of this study, burners were used to examine the repeatability of each fire's characteristics, the amount of heat transfered to a wall as a function of distance and the damage to a section of painted gypsum wallboard.  While the data generated from those full scale laboratory experiments provided valuable insight, it would be better to be able to develop assessments of fire patterns based on data from one or a series of bench scale tests in an effort to validate or add to the existing data in the literature.  The bench scale apparatus used in this study include: the cone calorimeter, the radiant flooring panel, and a vertical radiant panel.  A lateral ignition and flamespread test (LIFT) apparatus may have also provided useful results for this study, but an operational LIFT apparatus was not available at NIST over the course of this study.

@@

The cone calorimeter is one of the most versitile and commonly used bench-scale apparatus that uses the oxygen consumption calorimetry method to measure heat release rate. It was developed by Babrauskas at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted by the ASTM as the Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter, ASTM~E1354~\cite{ASTM_E1354}. The cone calorimeter consists of a conically shaped heater, shutter, spark ignitor, sample holder, and load cell located underneath an exhaust hood. In addition, the instrumentation required to measure mass flow, gas concentrations, and optical density are connected to the exhaust duct work.  Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/Cone_Cal} \\
	\caption[Schematic of the cone calorimeter]{Schematic of the cone calorimeter adapted from~\cite{Babrauskas:1992}}
	\label{Cone_Cal}
\end{figure}

For the ignition tests, an electric spark igniter was positioned over the center of the horizontal samples to serve as a pilot for the pyrolysis gases.  Combustion products and dilution air were extracted through the hood and exhaust duct by a high temperature fan. The flow rate may be adjusted between 0.01 and 0.03~m$^3$/s. The volumetric flow rate was kept constant during testing. The sample was mounted on a load cell to determine mass loss rate during a test, and smoke obscuration was measured using a laser light source. The gas flow rate in the exhaust duct was calculated from the pressure drop across and temperature at an orifice plate in the duct. The concentrations of oxygen, carbon dioxide, carbon monoxide, were measured using appropriate instruments. Heat release rate was calculated from the gas concentration and mass flow measurements.
